A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-08-06 manuscript/manuscript.Rmd
I made some edits to the start of the intro

---
title             : "A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data"
shorttitle        : "Tutorial on hazard analysis"

author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "ETH Zürich"

authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).

  <!-- Enter author note here. -->

abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. To move beyond mean performance measures, various distributional analyses have been proposed. Here we focus on one particular distributional analysis known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. Along the way, we discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time survival analysis.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "response times, event history analysis, Bayesian regression models"
wordcount         : "X"

bibliography      : ["r-references.bib", "extrareferences.bib"]

floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
# output            : papaja::apa6_word
editor_options: 
  chunk_output_type: console
header-includes:
  - \raggedbottom
---

@@

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations using analysis-of-variance. However, differences in means conceal when an experimental effect starts, how long it lasts, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].
In this tutorial we focus on a distributional method for analyzing time-to-event data that is known as discrete-time event history analysis (EHA), a.k.a. survival, hazard, duration, failure-time, and transition analysis [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @allisonSurvivalAnalysisUsing2010]. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4.

>>>>

---
title             : "A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data"
shorttitle        : "Tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. To move beyond mean performance measures, various distributional analyses have been proposed. Here we focus on one particular distributional analysis known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. Across four tutorials that we make publicly available on Github and OSF, we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. Along the way, we discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time survival analysis.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "response times, event history analysis, Bayesian regression models"
wordcount         : "X"
bibliography      : ["r-references.bib", "extrareferences.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
# output            : papaja::apa6_word
editor_options: 
  chunk_output_type: console
header-includes:
  - \raggedbottom
---

@@

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating mean average performance across a series of trials. However, differences in means conceal when an experimental effect starts, how long it lasts, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].
In this tutorial we focus on a distributional method for analyzing time-to-event data that is known as discrete-time event history analysis (EHA), a.k.a. survival, hazard, duration, failure-time, and transition analysis [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @allisonSurvivalAnalysisUsing2010]. Across four tutorials that we make publicly available on Github and the Open Science Framework (OSF), we illustrate how to calculate and interpret descriptive statistics, and how to implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4.
